---
output:
  pdf_document:
    highlight: null
fontsize: 10pt
geometry: margin=1.25in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Box 1 - Example DFA workflow

## Model specification

For this example we will assume that the observations at time $t$ ($\mathbf{y}_t$) arise from a multivariate normal distribution wherein the mean vector is the sum of a vector of intercepts ($\boldsymbol{\alpha}$) and latent factors ($\mathbf{L} \boldsymbol{\gamma}_t$), and the covariance matrix $\boldsymbol{\Phi}$ is diagonal and unequal. Eqn (1) then becomes:

$$
\mathbf{y}_t \sim \text{MVN}(\boldsymbol{\alpha} + \mathbf{L} \boldsymbol{\gamma}_t, \boldsymbol{\Phi}).
$$


The latent factors are themselves a multivariate random walk with normal errors. Here we assume $\mathbf{Q}$ equals the identity matrix $\mathbf{I}$, such that Eqn (3) is

$$
\boldsymbol{\gamma}_t \sim \text{MVN}(\boldsymbol{\gamma}_{t-1},\mathbf{Q}).
$$

## Simulated data

Here are some simulated data for 5 time series with 30 measurements each based upon 2 latent factors. The observation variances are also unequal. We will set the covariance of the process errors $\mathbf{Q}$ to the identity matrix $\mathbf{I}$. Note that we will also subtract the mean of the data so as to aid in parameter estimation (see below). 

\vspace{0.15in}

```{r sim_data, size="small"}
set.seed(123)
## matrix dims
nn <- 5; mm <- 2; tt <- 30
## zero-mean latent factors
gamma1 <- cumsum(rnorm(tt)); gamma1 <- gamma1 - mean(gamma1)
gamma2 <- cumsum(rnorm(tt)); gamma2 <- gamma2 - mean(gamma2)
## factor loadings
LL <- matrix(c(0.2, 0.9,
               0.4, 0.7,
               0.6, 0.5,
               0.8, 0.3,
               1.0, 0.1),
             nn, mm, byrow=TRUE)
## covariance matrix for observation errors
Phi <- diag(seq(1,5)/5)
## observation errors
vv <- t(MASS::mvrnorm(tt, matrix(0,nn,1), Phi))
## simulated data
yy <- LL %*% rbind(gamma1,gamma2) + vv
## zero-mean data
yy <- t(scale(t(yy), scale=FALSE))
```


## Fitting the model

There are several options for fitting factor models, but for this example we make use of the **MARSS** package (Holmes et al. 2012) for **R** to estimate the parameters in the DFA model. The notation of Holmes et al. is somewhat different than ours, however, so we must write out our DFA model in a form based on the following specification of a state-space model:

$$
\begin{gathered}
\mathbf{y}_t \sim \text{MVN}(\mathbf{a} + \mathbf{Z} \mathbf{x}_t, \mathbf{R}) \\
\mathbf{x}_t \sim \text{MVN}(\mathbf{u} + \mathbf{B} \mathbf{x}_{t-1}, \mathbf{Q}).
 \end{gathered}   
$$

Thus, for our DFA model we want $\mathbf{a} = \boldsymbol{\alpha}$, $\mathbf{Z} = \mathbf{L}$ $\mathbf{x}_t = \boldsymbol{\gamma}_t$, $\mathbf{R} = \boldsymbol{\Phi}$, $\mathbf{u} = \mathbf{0}$, and $\mathbf{B} = \mathbf{I}$. Furthermore, because we substracted the mean of each time series, we will set $\mathbf{a} = \mathbf{0}$. Note that this will not affect our estimates of the factors and their loadings.

The `MARSS` function makes use of list matrices, which allow one to mix character and numeric classes within the same matrix. Any entries of class `character` are interpreted as parameters to be estimated; any `numeric` entries are fixed at their value.

\vspace{0.15in}

```{r fit_dfa, size="small", results="hide"}
library(MARSS)
## model list
mod_list <- list(
  ## observation eqn
  A = matrix(0,nn,1),        # zero vector
  Z = matrix(list(0),nn,mm), # all 0's for now
  R = matrix(list(0),nn,nn), # all 0's for now
  ## state eqn
  U = matrix(0,mm,1),        # zero vector
  B = diag(mm),              # identity matrix
  Q = diag(mm)               # identity matrix
)
## specify observation parameters
## upper right corner of Z stays zero
for(cc in 1:mm) {
  for(rr in cc:nn) {
    mod_list$Z[rr,cc] <- paste0(rr,cc)
  }
}
## R is diagonal and unequal
diag(mod_list$R) <- paste0(seq(nn),seq(nn))
## fit the model
dfa <- MARSS(yy, model=mod_list, inits=list(x0=matrix(0,2,1)))
```

## Factor rotations

For this example we will use a varimax rotation of the factors and loadings, where we seek an $m \times m$ non-singular rotation matrix $\mathbf{H}$, such that the following state-space forms of our DFA model are equivalent ($\mathbf{v}_t$ and $\mathbf{w}_t$ are vectors of observation and process errors, respectively):

$$
\begin{aligned}
 \mathbf{y}_t = \mathbf{L} \boldsymbol{\gamma}_t + \boldsymbol{\alpha} +\mathbf{v}_t
 &\Leftrightarrow \mathbf{y}_t = \mathbf{L}\mathbf{H}^{-1} \boldsymbol{\gamma}_t + \boldsymbol{\alpha} + \mathbf{v}_t \\
 \boldsymbol{\gamma}_t = \boldsymbol{\gamma}_{t-1} + \mathbf{w}_t
 &\Leftrightarrow \mathbf{H} \boldsymbol{\gamma}_t = \mathbf{H} \boldsymbol{\gamma}_{t-1} + \mathbf{H} \mathbf{w}_t.
\end{aligned}
$$

This is easily done in **R** via the `varimax` function:

\vspace{0.15in}

```{r varimax, size="small"}
## loadings matrix
L_hat <- coef(dfa, type="matrix")$Z
## inverse of the rotation matrix
H_inv <- varimax(L_hat)$rotmat
## rotate loadings matrix
L_rot <- L_hat %*% H_inv
## rotate factors
gamma_rot <- solve(H_inv) %*% dfa$states
## model fits
yhat <- L_rot %*% gamma_rot
```

## Estimated parameters

Here are plots of the true (solid) and estimated factors (dashed). Note that **MARSS** has no way of ordering the factors, so care should be used when examining them. In this case, the estimated factors were interchanged.

\vspace{0.15in}

```{r plot_factors, size="small", fig.height=3.5, fig.width=5.5, fig.align="center"}
matplot(cbind(gamma2,gamma1,t(gamma_rot)), type="l", lty=c(1,1,2,2), lwd=2, 
        col=c("darkblue", "darkred"), ylab="Factors", xlab="Time")
```

Here is a comparison of the true factor loadings and the estimates in $\hat{\mathbf{L}}$.

\vspace{0.15in}

```{r fac_loadings, size="small"}
## true values
LL
## estimates
round(L_rot[,2:1],2)
```

It looks like the estimated loadings are generally pretty close to the true values, with the exception of the loading of factor 2 on data series 5. 

## Model fits

Here is a plot of the simulated data (numbers) and the fits (lines) from the estimated DFA model.

\vspace{0.15in}

```{r plot_fits, size="small", fig.height=3.5, fig.width=5.5, fig.align="center"}
matplot(t(yy), type="p", cex=0.8, col=rainbow(nn,start=0.55, end=0.85),
        ylab="Observations", xlab="Time")
matlines(t(yhat), type="l", lty=1, lwd=2, col=rainbow(nn,start=0.55, end=0.85))
```

